%%-----------Chapter 6------------------------------------------
\chapter{Work, Energy, and Momentum}

\section{Introduction: The Grand Conservation Laws}

In the previous chapters, we analyzed motion by looking at forces moment by moment. We asked: what is the acceleration right now? Where will the particle be in the next instant? This is Newton’s direct path, and it is incredibly powerful. But as systems grow from a single particle to many interacting pieces, tracking every force at every instant becomes a nightmare. 

Fortunately, nature has hidden shortcuts. If we step back and look at the beginning and the end of a process, we discover that certain quantities remain completely unchanged, no matter how chaotic the middle was. These are the **conservation laws**. In this chapter, we will meet three of them: the conservation of linear momentum, the conservation of energy, and the conservation of angular momentum. Together, they form the backbone of classical mechanics and provide a modern, first-principles way to understand everything from a colliding billiard ball to a launching rocket and an orbiting planet.

\section{Momentum and Its Conservation}

Let us begin with a simple observation. A cricket ball and a tennis ball are thrown at you at the same speed. Which is harder to catch? The cricket ball, of course. Now imagine a tennis ball thrown at a gentle pace versus one served by a professional at \SI{200}{km/h}. Again, the faster one is harder to catch.

What makes an object hard to stop is a combination of its mass and its velocity. Newton called this the ``quantity of motion.'' Today, we call it **linear momentum**, denoted by the vector $\vec{p}$.

\begin{keybox}[Linear Momentum]
The **linear momentum** $\vec{p}$ of a particle of mass $m$ moving with velocity $\vec{v}$ is the product of its mass and its velocity:
\begin{equation}
    \vec{p} = m \vec{v}.
\end{equation}
Momentum is a vector quantity. Its direction is the same as the velocity, and its SI unit is the kilogram-metre per second (\si{kg.m/s}).
\end{keybox}

Newton’s second law was originally written not as $\vec{F} = m\vec{a}$, but in terms of momentum:
\begin{equation}
    \vec{F} = \frac{d\vec{p}}{dt}. \label{eq:newton2_mom}
\end{equation}
If mass is constant, this recovers $\vec{F} = d(m\vec{v})/dt = m(d\vec{v}/dt) = m\vec{a}$. But Eq.~\eqref{eq:newton2_mom} is more fundamental because it holds even when mass is changing (like a rocket burning fuel) and in the relativistic domain where mass appears to increase with speed.

\subsection{Impulse and the Change in Momentum}

If we multiply a force by the short time interval $dt$ over which it acts, we get a quantity called **impulse**, $\vec{J}$:
\begin{equation}
    d\vec{J} = \vec{F}\,dt \implies \vec{J} = \int_{t_1}^{t_2} \vec{F}\,dt.
\end{equation}
Integrating Eq.~\eqref{eq:newton2_mom} from $t_1$ to $t_2$, we obtain the **impulse-momentum theorem**:
\begin{equation}
    \vec{J} = \vec{p}_2 - \vec{p}_1 = \Delta\vec{p}.
\end{equation}
The impulse applied to an object is exactly equal to its change in momentum. This explains why safety nets, crumple zones in cars, and catching a cricket ball with ``soft hands'' (pulling your hands back as you catch) reduce the impact force. By increasing the time interval $\Delta t$ over which the momentum change occurs, the average force is reduced: $\vec{F}_{\text{avg}} = \Delta\vec{p}/\Delta t$.

\begin{workedexample}[The Soft Hands of a Fielder]
A cricket ball of mass $m = \SI{0.16}{kg}$ reaches a fielder traveling at $v = \SI{20}{m/s}$.
\begin{enumerate}[label=(\alph*), itemsep=1pt]
  \item What is the initial momentum of the ball?
  \item If the fielder catches the ball by holding their hands rigid, bringing it to rest in $\Delta t = \SI{0.01}{s}$, what is the average force felt by their hands?
  \item If they pull their hands back during the catch, extending the time to $\Delta t = \SI{0.15}{s}$, what is the average force?
\end{enumerate}
\textbf{Solution:}
\begin{enumerate}[label=(\alph*)]
  \item $p_i = m v = (\SI{0.16}{kg})(\SI{20}{m/s}) = \SI{3.2}{kg.m/s}$.
  \item The change in momentum is $\Delta p = p_f - p_i = 0 - 3.2 = -\SI{3.2}{kg.m/s}$. The average force is:
        \begin{equation*}
            F_{\text{avg}} = \frac{\Delta p}{\Delta t} = \frac{-\SI{3.2}{kg.m/s}}{\SI{0.01}{s}} = -\SI{320}{N}.
        \end{equation*}
        This is about \SI{32}{kg} of weight slamming into the hands.
  \item With the extended time:
        \begin{equation*}
            F_{\text{avg}} = \frac{-\SI{3.2}{kg.m/s}}{\SI{0.15}{s}} \approx -\SI{21.3}{N}.
        \end{equation*}
        Pulling the hands back reduces the force by a factor of 15, making the catch safe and painless.
\end{enumerate}
\end{workedexample}

\subsection{Conservation of Linear Momentum}

Consider a system of two colliding particles, $A$ and $B$. During the collision, particle $A$ exerts a force $\vec{F}_{B \to A}$ on $B$, and by Newton's third law, $B$ exerts an equal and opposite force $\vec{F}_{A \to B} = -\vec{F}_{B \to A}$ on $A$. If no external forces act on the system:
\begin{equation}
    \frac{d\vec{p}_A}{dt} + \frac{d\vec{p}_B}{dt} = \vec{F}_{B \to A} + \vec{F}_{A \to B} = \vec{0} \implies \frac{d}{dt}(\vec{p}_A + \vec{p}_B) = \vec{0}.
\end{equation}
Thus, the total momentum $\vec{P}_{\text{total}} = \vec{p}_A + \vec{p}_B$ is constant.

\begin{keybox}[Conservation of Linear Momentum]
If the net external force acting on a system of particles is zero, the total linear momentum of the system remains constant:
\begin{equation}
    \vec{P}_{\text{initial}} = \vec{P}_{\text{final}}.
\end{equation}
\end{keybox}

\begin{predictbox}[6.1: The Recoil of a Rifle]
A heavy rifle of mass $M$ fires a light bullet of mass $m$ horizontally with velocity $\vec{v}_b$ relative to the ground. Before firing, everything is at rest. 
Predict: which direction will the rifle move when fired, and will its speed be greater than, equal to, or less than the bullet's speed? Commit to your prediction before reading further.
\end{predictbox}

Since the net external horizontal force on the rifle-bullet system is zero, momentum must be conserved:
\begin{equation}
    \vec{P}_i = \vec{0} \implies \vec{P}_f = M\vec{v}_r + m\vec{v}_b = \vec{0} \implies \vec{v}_r = -\frac{m}{M}\vec{v}_b.
\end{equation}
The negative sign means the rifle moves in the opposite direction (recoil). Since the rifle is much heavier than the bullet ($M \gg m$), its speed is much smaller ($v_r \ll v_b$).

\subsection{Rocket Propulsion: The Tsiolkovsky Equation}

A rocket is a beautiful application of momentum conservation in which the mass of the system changes continuously as fuel is burned and ejected.

Let a rocket of mass $m$ be moving with speed $v$ at time $t$. In a short time interval $dt$, it ejects a mass $dm$ of gas backward with an exhaust speed $u$ relative to the rocket (which means speed $v - u$ relative to the ground). The rocket's mass becomes $m - dm$, and its speed increases to $v + dv$. 

Applying momentum conservation in the ground frame (ignoring gravity for now):
\begin{equation}
    P(t) = P(t+dt) \implies m v = (m - dm)(v + dv) + dm(v - u).
\end{equation}
Expanding this and ignoring the tiny second-order product $dm\,dv$:
\begin{equation}
    m v = m v + m\,dv - v\,dm - u\,dm + v\,dm \implies m\,dv = u\,dm.
\end{equation}
Since the rocket's mass decreases, the change in the rocket's mass is $dm = -dm_{\text{rocket}}$. Let us write this as $m\,dv = -u\,dm$, where $dm$ is now the differential change in the rocket's own mass ($dm < 0$). Dividing by $m$ and integrating:
\begin{equation}
    \int_{v_0}^{v} dv' = -u \int_{m_0}^{m} \frac{dm'}{m'} \implies v - v_0 = u \ln\left(\frac{m_0}{m}\right).
\end{equation}
This is the famous **Tsiolkovsky rocket equation**, derived independently by the Russian schoolteacher Konstantin Tsiolkovsky in 1897. It shows that to reach high speeds, a rocket must either have an extremely high exhaust velocity $u$ or eject a massive fraction of its initial mass as fuel ($m_0/m \gg 1$).

\section{Work and Kinetic Energy}

We now shift to our second conservation law: energy. Historically, the concept of energy was much slower to develop than momentum. While Newton's mechanics focused on forces and momentum, Leibniz and others argued that another quantity, which they called \emph{vis viva} (``living force'') and computed as $mv^2$, was also conserved in certain collisions. This was the precursor to kinetic energy.

To connect forces to energy, we define the concept of **work**.

\subsection{Work Done by a Force}

If a constant force $\vec{F}$ acts on an object while it undergoes a displacement $\vec{d}$, the work $W$ done by the force is:
\begin{equation}
    W = \vec{F} \cdot \vec{d} = F d \cos\phi,
\end{equation}
where $\phi$ is the angle between the force and displacement vectors. If the force varies along the path, we must integrate:
\begin{equation}
    W = \int_{\vec{r}_1}^{\vec{r}_2} \vec{F} \cdot d\vec{r}.
\end{equation}
Work is a scalar quantity. Its SI unit is the newton-metre, which we rename the **joule** (\si{J}) in honour of James Prescott Joule.

\subsection{The Work-Kinetic Energy Theorem}

What does work do? Let us compute the work done by a net force $\vec{F}_{\text{net}}$ on a particle of mass $m$ as it moves along a path:
\begin{equation}
    W = \int_{1}^{2} \vec{F}_{\text{net}} \cdot d\vec{r} = \int_{1}^{2} m \frac{d\vec{v}}{dt} \cdot d\vec{r}.
\end{equation}
Since $\vec{v} = d\vec{r}/dt$, we can swap the variables using the chain rule: $d\vec{r} = \vec{v}\,dt$. Thus:
\begin{equation}
    W = \int_{t_1}^{t_2} m \frac{d\vec{v}}{dt} \cdot \vec{v}\,dt = \int_{v_1}^{v_2} m \vec{v} \cdot d\vec{v}.
\end{equation}
Since $\vec{v} \cdot d\vec{v} = v\,dv$ (the change in speed squared), we get:
\begin{equation}
    W = \int_{v_1}^{v_2} m v\,dv = \frac{1}{2}m v_2^2 - \frac{1}{2}m v_1^2.
\end{equation}

\begin{keybox}[Work-Kinetic Energy Theorem]
The net work done on a particle is exactly equal to the change in its **kinetic energy** $K$:
\begin{equation}
    W_{\text{net}} = \Delta K, \quad \text{where } K \equiv \frac{1}{2}m v^2.
\end{equation}
\end{keybox}

This theorem connects kinematic changes in speed directly to the forces acting over distances.

\section{Potential Energy and Conservative Forces}

Some forces, like gravity or a spring, are special: the work they do on an object as it moves from point $A$ to point $B$ depends only on the endpoints $A$ and $B$, not on the path taken. Such forces are called **conservative**.

\begin{keybox}[Conservative Forces]
A force $\vec{F}$ is **conservative** if either of these equivalent statements holds:
\begin{enumerate}
  \item The work done by $\vec{F}$ around any closed loop is zero: $\oint \vec{F} \cdot d\vec{r} = 0$.
  \item The work done by $\vec{F}$ between two points is independent of the path connecting them.
\end{enumerate}
\end{keybox}

For any conservative force, we can define a scalar function called **potential energy**, $U(\vec{r})$, such that the work done by the force is the negative change in this potential energy:
\begin{equation}
    W_{\text{cons}} = -\Delta U = U_1 - U_2.
\end{equation}
In one dimension, this relationship is:
\begin{equation}
    F_x = -\frac{dU}{dx}.
\end{equation}
For gravity near the Earth's surface, $U(y) = mgy$ (where $F_y = -mg$). For a spring obeying Hooke's law ($F_x = -kx$), $U(x) = \frac{1}{2}kx^2$.

\subsection{Conservation of Mechanical Energy}

If only conservative forces do work on a system, then $W_{\text{net}} = W_{\text{cons}} = -\Delta U$. But by the work-KE theorem, $W_{\text{net}} = \Delta K$. Equating these:
\begin{equation}
    \Delta K = -\Delta U \implies \Delta(K + U) = 0.
\end{equation}
We define the **total mechanical energy** as $E \equiv K + U$. Thus, if no non-conservative forces (like friction or drag) do work:
\begin{equation}
    E = K + U = \text{constant}.
\end{equation}

If non-conservative forces *do* work (like friction $f_k$), we write:
\begin{equation}
    W_{\text{net}} = W_{\text{cons}} + W_{\text{nc}} = -\Delta U + W_{\text{nc}} = \Delta K \implies E_2 - E_1 = W_{\text{nc}}.
\end{equation}
Since friction always does negative work, mechanical energy is lost to heat (thermal energy). The total energy of the universe is still conserved, but mechanical energy is dissipated.

\begin{workedexample}[The Loop-the-Loop]
A small block of mass $m$ slides down a frictionless track and enters a circular loop of radius $R$. What is the minimum height $h$ from which the block must start (from rest) to make it all the way around the loop without falling off at the top?
\newline
\textbf{Solution:}
Let us first look at the top of the loop (point $T$, height $2R$). Two forces act on the block: gravity $mg$ downward and the normal force $\mathscr{N}$ downward. For circular motion:
\begin{equation*}
    \Sigma F_{\text{radial}} = mg + \mathscr{N} = m \frac{v_T^2}{R}.
\end{equation*}
To just maintain contact, the normal force must not drop below zero ($\mathscr{N} \ge 0$). Thus, the minimum speed required at the top is:
\begin{equation*}
    v_T^2 \ge gR.
\end{equation*}
Now we use conservation of mechanical energy between the starting point (height $h$, speed $v_0 = 0$) and the top (height $2R$, speed $v_T$):
\begin{align*}
    E_{\text{initial}} &= E_{\text{final}} \\
    mgh + 0 &= mg(2R) + \frac{1}{2}m v_T^2.
\end{align*}
Canceling $m$ and substituting the threshold speed $v_T^2 = gR$:
\begin{equation*}
    gh = 2gR + \frac{1}{2}gR \implies h = \frac{5}{2}R.
\end{equation*}
To clear the loop, the block must start at a height at least $2.5$ times the radius of the loop.
\end{workedexample}

\begin{checkpoint}[6.2]
A cart is released from rest at height $h$ on a roller coaster track. It goes down a valley and up a hill of height $H < h$. If friction is present:
\newline
\begin{tabular}{ll}
  A. The speed at the top of the hill is $\sqrt{2g(h-H)}$. & B. The speed is less than $\sqrt{2g(h-H)}$. \\
  C. The speed is greater than $\sqrt{2g(h-H)}$. & D. The cart cannot reach the hill.
\end{tabular}
\checkanswer{B. Because friction does negative work, some mechanical energy is lost to thermal energy, so the speed at the top of the hill will be less than the frictionless prediction.}
\end{checkpoint}

\section{Collisions in 2D}

Let us bring momentum and energy together to analyze collisions. In any collision where the net external force is zero, **momentum is conserved**. However, kinetic energy is not always conserved:
- An **elastic collision** is one in which the total kinetic energy is conserved: $K_i = K_f$.
- An **inelastic collision** is one in which kinetic energy is not conserved: $K_i > K_f$ (energy is lost to sound, deformation, and heat). If the objects stick together, the collision is **perfectly inelastic**.

In two dimensions, we write momentum conservation as vector components:
\begin{align}
    P_{i,x} &= P_{f,x} \\
    P_{i,y} &= P_{f,y}.
\end{align}

\begin{workedexample}[2D Elastic Collision of Identical Balls]
A billiard ball $A$ of mass $m$ moving with speed $v$ along the $x$-axis collides elastically with an identical ball $B$ of mass $m$ at rest. After the collision, ball $A$ moves off at angle $\theta_A$ relative to the $x$-axis, and ball $B$ moves off at angle $\theta_B$.
Prove that the angle between their final paths, $\theta_A + \theta_B$, is exactly $90^\circ$ (unless the collision is head-on).
\newline
\textbf{Solution:}
Let $\vec{v}$ be the initial velocity of ball $A$. Let $\vec{v}_A$ and $\vec{v}_B$ be the final velocities of the two balls.
Conservation of linear momentum gives:
\begin{equation*}
    m\vec{v} = m\vec{v}_A + m\vec{v}_B \implies \vec{v} = \vec{v}_A + \vec{v}_B.
\end{equation*}
Let us take the dot product of this vector equation with itself:
\begin{equation}
    v^2 = (\vec{v}_A + \vec{v}_B) \cdot (\vec{v}_A + \vec{v}_B) = v_A^2 + v_B^2 + 2\vec{v}_A \cdot \vec{v}_B. \label{eq:coll_dot}
\end{equation}
Because the collision is elastic, kinetic energy is also conserved:
\begin{equation*}
    \frac{1}{2}mv^2 = \frac{1}{2}mv_A^2 + \frac{1}{2}mv_B^2 \implies v^2 = v_A^2 + v_B^2.
\end{equation*}
Subtracting this from Eq.~\eqref{eq:coll_dot}, we find:
\begin{equation*}
    2\vec{v}_A \cdot \vec{v}_B = 0 \implies \vec{v}_A \cdot \vec{v}_B = 0.
\end{equation*}
If the dot product of the final velocity vectors is zero, they must be perpendicular! Thus, the angle between them is $\theta_A + \theta_B = 90^\circ$. (This holds for any non-head-on elastic collision of identical masses).
\end{workedexample}

\section{The Center of Mass}

If we drop a spinning, vibrating wrench, its motion looks incredibly complex. But if we paint a bright dot at one specific point on the wrench, we discover that the dot traces out a perfect parabola, exactly like a single point particle. That special point is the **center of mass**.

\begin{keybox}[Center of Mass]
For a system of $N$ discrete particles with masses $m_i$ and positions $\vec{r}_i$, the position of the center of mass $\vec{R}_{\text{cm}}$ is:
\begin{equation}
    \vec{R}_{\text{cm}} = \frac{1}{M}\sum_{i=1}^{N} m_i \vec{r}_i, \quad \text{where } M \equiv \sum_{i=1}^{N} m_i.
\end{equation}
For a continuous object of mass density $\rho$, we integrate:
\begin{equation}
    \vec{R}_{\text{cm}} = \frac{1}{M}\int \vec{r}\,dm.
\end{equation}
\end{keybox}

Differentiating the position of the center of mass with respect to time, we find:
\begin{equation}
    \vec{V}_{\text{cm}} = \frac{d\vec{R}_{\text{cm}}}{dt} = \frac{1}{M}\sum m_i \vec{v}_i = \frac{\vec{P}_{\text{total}}}{M}.
\end{equation}
Thus, the total momentum of a system is simply the total mass multiplied by the velocity of the center of mass: $\vec{P}_{\text{total}} = M \vec{V}_{\text{cm}}$. If no net external force acts, $\vec{P}_{\text{total}}$ is constant, which means the center of mass moves at constant velocity, completely unaffected by any internal collisions or explosions!

\section{Angular Momentum and Torque}

Just as linear momentum describes an object's resistance to stopping its straight-line motion, **angular momentum** describes its resistance to stopping its rotation.

For a single particle of mass $m$ at position $\vec{r}$ relative to an origin, moving with momentum $\vec{p} = m\vec{v}$, we define its angular momentum $\vec{L}$ relative to that origin using the cross product:
\begin{keybox}[Angular Momentum of a Particle]
The **angular momentum** $\vec{L}$ of a particle is:
\begin{equation}
    \vec{L} \equiv \vec{r} \times \vec{p}.
\end{equation}
Its magnitude is $L = r p \sin\theta$, where $\theta$ is the angle between $\vec{r}$ and $\vec{p}$. The units of angular momentum are \si{kg.m^2/s}.
\end{keybox}

What changes angular momentum? Let us take the time derivative of $\vec{L}$:
\begin{equation}
    \frac{d\vec{L}}{dt} = \frac{d}{dt}(\vec{r} \times \vec{p}) = \left(\frac{d\vec{r}}{dt} \times \vec{p}\right) + \left(\vec{r} \times \frac{d\vec{p}}{dt}\right).
\end{equation}
The first term is $\vec{v} \times (m\vec{v}) = m(\vec{v} \times \vec{v}) = \vec{0}$, since the cross product of any vector with itself is zero. The second term is $\vec{r} \times \vec{F}_{\text{net}}$, using Newton's second law. We call the cross product of position and force the **torque**, $\vec{\tau}$:
\begin{equation}
    \vec{\tau} \equiv \vec{r} \times \vec{F}.
\end{equation}
Thus, we arrive at the rotational version of Newton's second law:
\begin{equation}
    \vec{\tau}_{\text{net}} = \frac{d\vec{L}}{dt}.
\end{equation}
If the net torque on a system is zero, its angular momentum is conserved: $\vec{L}_{\text{initial}} = \vec{L}_{\text{final}}$.

\subsection{Kepler's Second Law: A Disguised Conservation Law}

We promised in Chapter 5 that Kepler's second law --- that a planet's position vector sweeps out equal areas in equal times --- is actually just the conservation of angular momentum in disguise. Let us prove it.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=1.2]
  % Sun
  \draw[fill=orange!80, orange] (0,0) circle (0.15);
  \node[below] at (0,-0.18) {\small Sun};
  
  % Orbit path segment
  \draw[dashed, thin] (-0.5,1.8) to[out=10, in=180] (2,1.5) to[out=0, in=120] (4,0.5);
  
  % Planet positions
  \coordinate (P1) at (2,1.5);
  \coordinate (P2) at (2.8,1.4);
  
  % vectors
  \draw[-latex, myblue, thick] (0,0) -- (P1) node[midway, left] {$\vec{r}$};
  \draw[-latex, myred, thick] (P1) -- (P2) node[midway, above] {$d\vec{r}$};
  \draw[myblue, thin] (0,0) -- (P2);
  
  % swept area shading
  \fill[myblue!15, opacity=0.5] (0,0) -- (P1) -- (P2) -- cycle;
  
  \node[above right] at (2.8,1.4) {\small Planet};
  \node at (1.2,0.8) {$dA$};
\end{tikzpicture}
\caption{The area $dA$ swept out by a planet's radius vector $\vec{r}$ in time $dt$ is a triangle defined by $\vec{r}$ and the displacement $d\vec{r}$.}
\label{fig:kepler_proof}
\end{figure}

The area of the tiny triangle swept out by the radius vector $\vec{r}$ as the planet moves through displacement $d\vec{r} = \vec{v}\,dt$ is half the area of the parallelogram formed by $\vec{r}$ and $d\vec{r}$:
\begin{equation}
    dA = \frac{1}{2} |\vec{r} \times d\vec{r}| = \frac{1}{2} |\vec{r} \times \vec{v}|\,dt.
\end{equation}
Dividing by $dt$, the rate at which area is swept out (the areal velocity) is:
\begin{equation}
    \frac{dA}{dt} = \frac{1}{2} |\vec{r} \times \vec{v}|.
\end{equation}
Since $\vec{p} = m\vec{v}$, we can write $\vec{r} \times \vec{v} = (\vec{r} \times \vec{p})/m = \vec{L}/m$. Thus:
\begin{equation}
    \frac{dA}{dt} = \frac{|\vec{L}|}{2m}. \label{eq:areal_vel}
\end{equation}
Now, what is the torque acting on the planet? Gravity points directly along the line connecting the planet and the Sun. Thus, the force $\vec{F}_g$ is parallel to the radius vector $\vec{r}$. The torque is:
\begin{equation}
    \vec{\tau} = \vec{r} \times \vec{F}_g = \vec{0}.
\end{equation}
Since the torque is zero, the planet's angular momentum $\vec{L}$ is constant! Since $L$ and the mass $m$ are constant, Eq.~\eqref{eq:areal_vel} tells us that $dA/dt$ is constant. 

The planet sweeps out equal areas in equal times because gravity is a **central force** (pointing directly toward the source), which ensures that the planet's angular momentum remains perfectly conserved. What seemed a wonder is no wonder.

\begin{whiteboard}[6.1: Spinning in a Chair]
Sit in a swivel office chair that spins freely. Hold two heavy books in your hands and stretch your arms out wide. Have a partner spin you gently. Now, pull your arms in tight to your chest.
\begin{enumerate}[label=(\alph*), leftmargin=2.5em, itemsep=1pt]
  \item Predict: what will happen to your spinning speed when you pull your arms in?
  \item Why? Use the conservation of angular momentum. Identify what happens to your distribution of mass (moment of inertia) and how that affects your angular speed $\omega$.
  \item If you extend your arms again, does your speed return to its original value?
\end{enumerate}
\end{whiteboard}

\section{Summary of Conservation Laws}

\begin{formulabox}[The Three Pillars of Mechanics]
\begin{itemize}
  \item **Linear Momentum**: $\vec{p} = m\vec{v}$. Conserved if net external force is zero: $\Sigma \vec{F}_{\text{ext}} = \vec{0}$.
  \item **Mechanical Energy**: $E = K + U$, where $K = \frac{1}{2}mv^2$. Conserved if only conservative forces do work: $W_{\text{nc}} = 0$.
  \item **Angular Momentum**: $\vec{L} = \vec{r} \times \vec{p}$. Conserved if net external torque is zero: $\Sigma \vec{\tau}_{\text{ext}} = \vec{0}$.
\end{itemize}
\end{formulabox}